Framing active school travel in Ontario, or how spinach is good for you

### 2024-05-07 AST-Framing-Ontario/AST-Framing-Ontario.Rmd
Added package with files for Elisabetta

---
title: "Framing active school travel in Ontario, or how spinach is good for you"
author:
  - name: Elise Desjardins
    email: desjae@mcmaster.ca
    affiliation: McMaster University
  - name: Jason Lam
    email: lamj54@mcmaster.ca
    affiliation: McMaster University
  - name: Darcy Reynard 
    email: reynard@ualberta.ca
    affiliation: University of Alberta
  - name: Damian Collins
    email: damian.collins@ualberta.ca
    affiliation: University of Alberta
  - name: E. Owen D. Waygood
    email: owen.waygood@polymtl.ca
    affiliation: Polytechnique Montreal
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    correspondingauthor: true
    footnote: 1
address:
  - code: McMaster University
    address: School of Earth, Environment and Society, 1280 Main St. W, Hamilton, Ontario, Canada L8S 4K1
  - code: University of Alberta
    address: Department of Earth and Atmospheric Sciences, 1-26 Earth Sciences Building, Edmonton, Alberta, Canada T6G 2E3
  - code: Polytechnique Montreal
    address: Department of Civil, Geological and Mining Engineering, 2500, chemin de Polytechnique Montreal, Montreal, Quebec, Canada H3T 1J4
footnote:
  - code: 1
    text: "Corresponding author."
abstract: |
  Active school travel (AST) is promoted in many jurisdictions, including Ontario, Canada, where there is provincial support for school travel planning (STP) efforts. Two pillars of school travel plans in the province are education and encouragement. For these pillars to stand, relevant stakeholders must be cognizant of and understand the issues and stakes. For this reason, we deem it important to understand how AST is communicated. In this research we adopt framing analysis to investigate the ways in which various organizations communicate to the public around AST. A frame is a central organizing idea or story line that provides meaning to a particular phenomenon, and thus helps to set the parameters for conversations about policy. In the case of AST, framing can influence what policy alternatives are perceived as available by children, parents, and their wider communities. The research is supported by natural language processing techniques applied to publicly available documents from Ontario stakeholders involved in school travel planning. We then compare the findings from these documents to a selection of academic studies on AST. We conclude that framing of AST in Ontario is mostly empirical-scientific in style, and largely in agreement with academic research on AST. However, it is essentially conservative and does not challenge the status quo of motorized travel. Furthermore, the frames tend to download responsibility for change to households, which limits the scope of policy alternatives by keeping collective action out of the frame.

@@

```{r load-packages, include=FALSE}
library(DiagrammeR)
library(dplyr)
library(ggraph)
library(ggplot2)
library(igraph)
library(kableExtra)
library(pdftools)
library(readr)
library(reshape2)
library(stringr)
library(text2vec)
library(textdata)
library(tidyr)
library(tidytext)
library(tm) 
library(tools)
library(topicmodels)
library(widyr)
library(word2vec)
library(wordcloud)
library(ldatuning)
library(quanteda)
```

@@

Our research is supported by natural language processing techniques, applied to the analysis of documents produced by three STP stakeholder groups in Ontario (municipalities, school boards, and transportation consortia). We assemble a corpus of texts from public sources that present information for the general public or parents who are interested in AST and examine word frequency, bigrams, and concordances in these selected documents. Furthermore, we use topic analysis to identify themes presented by each stakeholder group. We then compare the findings from these documents to a selection of studies on AST and explore the extent to which there is concordance between the scholarly literature on AST and materials shared with the public. The results of this analysis, along with manual checks of the corpus by the authors, lead us to conclude that AST is framed as a matter of individual choices instead of collective action. The message is reasonable but inherently conservative and it fails to communicate a sense of urgency about health and environmental issues, which may help to explain the modest success of related policies [@buliungSchoolTravelPlanning2011; @mammenSchoolTravelPlanning2014; @mammenActiveSchoolTravel2014].

@@

The relevance of policy frames with respect to education and encouragement is highlighted by @beland2014developing, who remarks "framing processes as public relations are central to policy issues related to sustainable transportation". Seeing how influential parental perceptions are to AST, it is important to understand AST-related communications, since the construction of frames for policy positions or public issues can activate or restrict a particular response in the intended audience [@panFramingAnalysisApproach1993]. Besides parents and children, framing can also affect broader policy support for AST in the community. Municipal representatives are perceived to be instrumental but the involvement of other stakeholder groups (e.g., busing consortia representatives and local residents) can be lacking [@buttazzoniSupportingActiveSchool2018]. In this sense, framing can be used to position existing solutions as suitable to address particular issues [@mahFramecriticalPolicyAnalysis2014], which may prevent the public from being aware of other policy approaches that challenge the status quo. Contrariwise, framing can be instrumental to revealing more diverse policy options than conventional wisdom would have [@bosomworth2015climate]. Ultimately, it is important to understand framing of policy issues because creating meaning through frames contributes to the constant process of shaping our collective understanding of issues, uncovering what we collectively find (un)acceptable and why [@beland2014developing].

@@

Both frames present AST in a positive light, by highlighting the potential benefits. However, neither frame appears to explain why declining rates of AST are a problem or convey any urgency to this issue so that it attracts the attention of parents, the general public, or policy makers. For example, parents who drive their children to school have reported concerns about traffic volume around schools [@mammenUnderstandingDriveEscort2012], but may not recognize that their own behavior contributes to the problem that is perceived to prevent their child from safely walking or bicycling to school [@collinsSafeJourneysEnterprising2001; @rothmanSchoolEnvironmentStudent2017]. There is evidence that negative framing is more effective at eliciting responses than positive framing [e.g., @kahneman1979interpretation; @waygoodCO2ValenceFraming2018]. In this case, the frames emphasize the positive side of active travel but leave the negative impacts of behaviors such driving children to school out of the frame.

@@

Overall, we find that the proposed solutions in the secondary frame reinforce some of the same ideas: 1) behavior change from households making different travel decisions (individual responsibility); 2) policies that create resources for AST (mainly through education and planning); and 3) engineering solutions like bicycle lanes. This reflects findings from the AST literature that a range of solutions are needed to address different factors that influence AST [_inter alia, see_ @mitraIndependentMobilityMode2013; @panterAttitudesSocialSupport2010]. The recognition of engineering changes may reflect the strong engagement of engineering staff and municipal representatives in the Canadian STP process [@buttazzoniSupportingActiveSchool2018; @mammenPuttingSchoolTravel2015], as well as the evidence that the built environment influences mode choice to school. However, these measures tend to support the desired behavior without confronting the problem behavior with interventions such as forbidding drop-off within 100 m of the school, restricting vehicular traffic to only local traffic around schools at school-commute times, such as found in countries where nearly no children die while going to school [@waygood2020japan].

@@

## Policy implications

Starting in 2017, the province of Ontario has had a policy of supporting communities develop Active School Travel initiatives. By 2021, this initiative had awarded millions of dollars to over 25 projects across the province, including for School Travel Plans (STP). As we noted in Section \ref{stp-frames}, STP in Ontario rely heavily on education and encouragement. From the get-go, these are not the strongest policy tools, compared to, say, construction of infrastructure, regulation, or transfers [see @ohare1989typology].

Given the weight that education and encouragement are expected to carry in the process, it is important to understand how the relevant issues are communicated to the public. The results of the research indicate that various authorities in Ontario frame AST in a conservative way that dilutes the force of STP interventions. If we had to paraphrase the way AST is framed across the province, it would be like this: "eating spinach is good for you; here are some tips to help you choose spinach." The issues are presented as a matter of personal choice, where action is purely optional, and lacking a sense of urgency. While emphasizing the risks may seem counterproductive, generating awareness of consequences of a lack of action has been found to be effective at motivating change [e.g., @lawlor2018risk; @waygoodCO2ValenceFraming2018]. Accordingly, communications with the public would do well to emphasize the risks associated with continued use of nonactive modes to school.

@@

Assuming that School Travel Plans will continue to be emphasized as one of the key policies to increase Active School Travel in Ontario, we suggest that the way the issues are framed by authorities and communities be revised as follows:

1. Active School Travel is already presented in a positive light ("it is good for you"), but the risks of inaction are not sufficiently communicated ("continued used of fossil fuels threatens our natural environment").
1. Individual choices are important ("here are tips to choose spinach"), but collective action is more effective ("we all need to eat some spinach"). Accordingly, AST should be presented as a choice that not only some individuals make, but importantly, that communities make _sometimes through their authorities_.
1. Recognize variations in capabilities in STP; framing AST as a war on some mobility styles confounds the need for certain mobility styles by people with various financial and physical capabilities.
1. Recognize the agency of children, as their voices do not seem to be properly reflected in any of the frames.

>>>>

---
title: "Framing active school travel in Ontario, or how spinach is good for you"
author:
  - name: Elise Desjardins
    email: desjae@mcmaster.ca
    affiliation: McMaster University
  - name: Jason Lam
    email: lamj54@mcmaster.ca
    affiliation: McMaster University
  - name: Darcy Reynard 
    email: reynard@ualberta.ca
    affiliation: University of Alberta
  - name: Damian Collins
    email: damian.collins@ualberta.ca
    affiliation: University of Alberta
  - name: E. Owen D. Waygood
    email: owen.waygood@polymtl.ca
    affiliation: Polytechnique Montreal
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    correspondingauthor: true
    footnote: 1
address:
  - code: McMaster University
    address: School of Earth, Environment and Society, 1280 Main St. W, Hamilton, Ontario, Canada L8S 4K1
  - code: University of Alberta
    address: Department of Earth and Atmospheric Sciences, 1-26 Earth Sciences Building, Edmonton, Alberta, Canada T6G 2E3
  - code: Polytechnique Montreal
    address: Department of Civil, Geological and Mining Engineering, 2500, chemin de Polytechnique Montreal, Montreal, Quebec, Canada H3T 1J4
footnote:
  - code: 1
    text: "Corresponding author."
abstract: |
  Active school travel (AST) is promoted in many jurisdictions, including Ontario, Canada, where there is provincial support for school travel planning (STP) efforts. Two pillars of school travel plans in the province are education and encouragement. For these pillars to stand, relevant stakeholders must be cognizant of and understand the issues and stakes. For this reason, we deem it important to understand how AST is communicated. In this research we adopt framing analysis to investigate the ways in which various organizations communicate to the public around AST. A frame is a central organizing idea or story line that provides meaning to a particular phenomenon, and thus helps to set the parameters for conversations about policy. In the case of AST, framing can influence what policy alternatives are perceived as available by children, parents, and their wider communities. The research is supported by natural language processing techniques applied to publicly available documents from Ontario stakeholders involved in school travel planning. We then compare the findings from these documents to a selection of academic studies on AST. We conclude that framing of AST in Ontario is mostly empirical-scientific in style, and largely in agreement with academic research on AST. However, it is essentially conservative and does challenge to the status quo of motorized travel. Furthermore, the frames tend to download responsibility for change to households, which limits the scope of policy alternatives by keeping collective action out of the frame.

@@

```{r load-packages, include=FALSE}
library(DiagrammeR) # Graph/Network Visualization
library(dplyr) # A Grammar of Data Manipulation
library(ggraph) # An Implementation of Grammar of Graphics for Graphs and Networks
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(igraph) # Network Analysis and Visualization
library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
library(pdftools) # Text Extraction, Rendering and Converting of PDF Documents
library(readr) # Read Rectangular Text Data
library(reshape2) # Flexibly Reshape Data: A Reboot of the Reshape Package
library(stringr) # Simple, Consistent Wrappers for Common String Operations
library(text2vec) # Modern Text Mining Framework for R
library(textdata) # Download and Load Various Text Datasets
library(tidyr) # Tidy Messy Data
library(tidytext) # Text Mining using 'dplyr', 'ggplot2', and Other Tidy Tools
library(tm) # Text Mining Package 
library(tools) # Tools for Package Development
library(topicmodels) # Topic Models
library(widyr) # Widen, Process, then Re-Tidy Data
library(word2vec) # Distributed Representations of Words
library(wordcloud) # Word Clouds
library(ldatuning) # Tuning of the Latent Dirichlet Allocation Models Parameters
library(quanteda) # Quantitative Analysis of Textual Data
```

@@

Our research is supported by natural language processing techniques, applied to the analysis of documents produced by three STP stakeholder groups in Ontario (municipalities, school boards, and transportation consortia). We assemble a corpus of texts from public sources that present information for the general public or parents who are interested in AST and examine word frequency, bigrams, and concordances in these selected documents. Furthermore, we use topic analysis to identify themes presented by each stakeholder group. We then compare the findings from these documents to a selection of studies on AST and explore the extent to which there is concordance between the scholarly literature on AST and materials shared with the public. The results of this analysis, along with manual checks of the corpus by the authors, lead us to conclude that AST is framed as a matter of individual choices instead of collective action. The message is reasonable but inherently conservative and it fails to communicate a sense of urgency about health and environmental issues, which may help to explain the modest success of related policies [@buliungSchoolTravelPlanning2011; @mammenSchoolTravelPlanning2014; @mammenActiveSchoolTravel2014]

@@

The relevance of policy frames with respect to education and encouragement is highlighted by @beland2014developing, who remarks "framing processes as public relations are central to policy issues related to sustainable transportation". Seeing how influential parental perceptions are to AST, it is important to understand AST-related communications, since the construction of frames for policy positions or public issues can activate or restrict a particular response in the intended audience [@panFramingAnalysisApproach1993]. Besides parents and children, framing can also affect broader policy support for AST in the community. Municipal representatives are perceived to be instrumental but the involvement of other stakeholder groups (e.g., busing consortia representatives and local residents) can be lacking [@buttazzoniSupportingActiveSchool2018]. In this sense, framing can be used to position existing solutions as suitable to address particular issues [@mahFramecriticalPolicyAnalysis2014], which may prevent the public from being aware of other policy approaches that challenge the status quo. Contrariwise, framing can be instrumental to revealing more diverse policy options than conventional wisdom would have [@bosomworth2015climate]. Ultimately, it is important to understand framing of policy issues because creating meaning through frames contributes to the constant process of shaping of our collective understanding of issues, uncovering what we collectively find (un)acceptable and why [@beland2014developing].

@@

Both frames present AST in a positive light, by highlighting the potential benefits. However, neither frame appears to explain why declining rates of AST are a problem or convey any urgency to this issue so that it attracts the attention of parents, the general public, or policy makers. For example, parents who drive their children to school have reported concerns about traffic volume around schools [@mammenUnderstandingDriveEscort2012], but may not recognize that their own behavior contributes to the problem that is perceived to prevent their child from safely walking or bicycling to school [@collinsSafeJourneysEnterprising2001; @rothmanSchoolEnvironmentStudent2017]. There is evidence that negative framing is more effective at eliciting responses than positive framing (e.g., Kahneman and Tversky, 1979; Waygood and Avineri, 2018). In this case, the frames emphasize the positive side of active travel but leave the negative impacts of behaviors such driving children to school out of the frame.

@@

Overall, we find that the proposed solutions in the secondary frame reinforce some of the same ideas: 1) behavior change from households making different travel decisions (individual responsibility); 2) policies that create resources for AST (mainly through education and plannning); and 3) engineering solutions like bicycle lanes. This reflects findings from the AST literature that a range of solutions are needed to address different factors that influence AST [_inter alia, see_ @mitraIndependentMobilityMode2013; @panterAttitudesSocialSupport2010]. The recognition of engineering changes may reflect the strong engagement of engineering staff and municipal representatives in the Canadian STP process [@buttazzoniSupportingActiveSchool2018; @mammenPuttingSchoolTravel2015], as well as the evidence that the built environment influences mode choice to school. However, these measures tend to support the desired behavior without confronting the problem behavior with interventions such as forbidding drop-off within 100 m of the school, restricting vehicular traffic to only local traffic around schools at school-commute times, such as found in countries where nearly no children die while going to school [@waygood2020japan].

@@

## Implications for school travel planning

As we noted in Section \ref{stp-frames} above, STP in Ontario relies heavily on education and encouragement. From the get-go, these are not the strongest tools of government action [compared to, say, regulation and transfers; see @ohare1989typology]. The conservative framing used further dilutes the force of STP interventions. If we had to paraphrase the way AST is framed across the province, it would be like this: "eating spinach is good for you; here are some tips to help you choose spinach." The issues are presented as a matter of personal choice, purely optional, and without a sense of urgency. While emphasizing the risks may seem counterproductive, this approach has been found to be effective at motivating change [e.g., @lawlor2018risk; @waygoodCO2ValenceFraming2018]. Accordingly, communications with the public could emphasize the risks associated with continued use of nonactive modes to school.

@@
